Navigating to a greener Europe with 24/7 hourly clean electricity procurement

### 2023-10-02 manuscript/sections/results.tex
results: complete section 3.5

\subsection{Uncovering mechanisms how 24/7 CFE commitments decarbonise local grids}
\label{subsec:mechanisms}

\vspace{10pt}
\begin{res}
    24/7 CFE procurement reduces system-level \co emissions via two mechanisms that can be isolated as profile and volume effects.
\end{res}

\vspace{5pt}
\begin{res}
    The difference in impact that 24/7 CFE procurement and 100\% annual matching policies have on system decarbonisation amplifies with a cleaner state of local grids.
\end{res}

\vspace{5pt}
\begin{res}
    24/7 CFE procurement reduces the needs for flexibility in the rest of the system.
\end{res}

As illustrated in Figures \ref{fig:10-2025-IE-p1-zone_emissions}, \ref{fig:10-2025-DE-p3-zone_emissions}, \ref{fig:10-2025-PL-p3-zone_emissions} and \ref{fig:10-2030-IE-p1-zone_emissions}, voluntary commitments to 24/7 clean energy procurement result in a considerable reduction of \co emissions in local electricity grids. 
In this section, we delve deeper into the underlying mechanisms responsible for these observed reductions aiming to enhance our understanding of the complex interplay between clean energy procurement and emissions reduction at the system level.

The reduction of system-wide emissions can be attributed to two primary mechanism driven by the dynamics of clean energy procurement and grid-supplied generation.

First, the timing mechanism, concerns the alignment of clean electricity supply with electricity demand of participating consumers. When a share of electricity consumption in a given region is closely alligned with CFE supply, the need for backup dispatchable generation is reduced. This effect can reduce utilization of fossil-based peak generators, such as \gls{ocgt} power plants, particularly during periods when wind and solar resources are scarce.

Second, the volume mechanism, pertains to the overall quantity of clean electricity generated by procured resources. 
The excess clean electricity sold to the grid can displace the emitting grid generators or reduce the need of electricity imports from the neighbouring regions.
The decarbonisation effect depends on the composition of the 24/7 portfolio, the volume of excess generation that is sold to the grid and the generation (and import) mix in the local zone when the excess electricity is sold.

In the following, we do an experiment to isolate the volume from the profile mechanism. 
We aim to attribute the reduction of \co emissions in a local zone either to the excess electricity or to the better alignment of procured generation with \gls{ci} consumers' demand.
In the experiment, we rerun the optimization with setting the rigt-hand side of the constraint \textit{TODO CORRECT REF} to zero, so that only the profile mechanism is present. The difference in emissions to the case with excess allowed, we assign to the volume mechanism.

\cref{fig:10-profile-volume.pdf} shows the results of the experiment -- the reduction of emissions in a local zone in Mt\co per year if 10\% of \gls{ci} consumers commit to 24/7 hourly procurement with 100\% CFE target with isolated contributions of the profile and volume mechanisms.
In other words, \cref{fig:10-profile-volume.pdf} attributes the difference between emissions in the \textit{reference} and the \textit{100\%} 24/7 CFE cases presented  Figures \ref{fig:10-2025-IE-p1-zone_emissions}, \ref{fig:10-2025-DE-p3-zone_emissions}, \ref{fig:10-2025-PL-p3-zone_emissions} and \ref{fig:10-2030-IE-p1-zone_emissions} (right columns) to the two decarbonisation mechanisms. 
The x-axis of the \cref{fig:10-profile-volume.pdf} lists different states of electricity grids where consumers following 24/7 approach can be located: a combination of four zones and two modelled years yields eight 'background' systems.

Two observations in (\cref{fig:10-profile-volume.pdf}) follow intuition: (i) the impact on local zone emissions of 24/7 hourly CFE procurement policy is notably different in absolute numbers (Mt\co per year) depending on \gls{ci} consumers' location, and (ii) the decarbonisation effect of voluntary clean electricity procurement reduces over time since electricity mix in all modelled zones becomes cleaner due to tightening of \gls{necp}s and decomissioning of old emitting power plant fleet. 
Isolation of the profile and volume mechanisms offers news insights. First, both mechanisms contribute to the system decarbonisation. Second, the cases of Germany and Poland 2030 show that the impact of both mechanisms is also present with the cleaner states of the grid in 2030.

The economic intuition suggests that self-cannibalisation effect of non-dispatchable generators, such as wind and solar, can decrease the impact of clean energy procurement on system-level decarbonisation.  
\cref{fig:10-hourly-annual.pdf} provides a nuanced view on this point.
The figure presents the \textit{additional} reduction of \co emissions in a local zone driven by the 100\% 24/7 hourly matching with clean electricity over the 100\% annual (volumetric) matching policy.
In other words, \cref{fig:10-hourly-annual.pdf} illustrates proportional impact of the emissions reduction acvieved by the \textit{RES100\%} and the \textit{100\%} 24/7 CFE policies presented in Figures \ref{fig:10-2025-IE-p1-zone_emissions}, \ref{fig:10-2025-DE-p3-zone_emissions}, \ref{fig:10-2025-PL-p3-zone_emissions} and \ref{fig:10-2030-IE-p1-zone_emissions} (left and right columns), and gives an intuition how the impact changes between 2025 and 2030.

A prominent observation in \cref{fig:10-hourly-annual.pdf} is that 100\% annual matching has a proportionally smaller effect on a system decarbonisation than 24/7 hourly matching in 2030 compared to 2025. 
This result is driven by the diminishing value of each additional megawatt of capacity of non-dispatchable generators, i.e., self-cannibalisation of solar and wind generators. 
The 24/7 hourly CFE matching have a stronger decarbonisation impact in a cleaner systems due to (i) the profile mechanism, (ii) the volume mechanism, since the procured resources have a more diverse portfolio including battery storage, long-duration energy storage or advanced clean firm technologies that allow for having clean electricity available in periods when system-wide wind and solar generation is scarce.

The decarbonisation mechanisms discussed above suggest a displacement of fossil-fueled generators in the rest of the system in presense of voluntary clean electricity procurement.
We illustrate this effect in \cref{fig:10-2030-DE-p1-system_capacity_diff.pdf}, which shows the difference in the generation capacity expansion of the whole European electricity system (excluding resources procured by participating customers) with and without commitments to clean electrity procurement (an illustraive scenario for Germany, 2030, technological palette~1). 
Hourly CFE procurement benefits the system by having a pronounced effect of flexibility needs reduction -- the system requires less battery storage and less peaker capacity in a form of \gls{ocgt}. 
In the selected example, 3.8~GW of load (10\% of \gls{ci} sectors in Germany) participate in the procurement.
In a case of hourly matching with 100\% CFE target, this reduces \gls{ocgt} investments by 4.6~GW and battery storage by 2.6~GW in the rest of the system.
The substitution of gas-fired capacity with clean resources procured by \gls{ci} consumers facilitates decarbonization of the entire system, as shown above.

\begin{figure}
    \centering
    \begin{subfigure}[t]{0.95\columnwidth}
        \centering
        \caption{Emissions reduction in a local zone by 24/7 hourly matching with 100\% target -- profile and volume effects isolated.}
        \includegraphics[width=0.95\columnwidth]{profile-volume.pdf}
        \label{fig:10-profile-volume.pdf}
    \end{subfigure}
    \begin{subfigure}[t]{0.95\columnwidth}
        \centering
        \caption{Additional emissions reduction in local zone by 100\% hourly over annual matching as percentage points of reference zone emissions.}
        \includegraphics[width=0.95\columnwidth]{hourly-annual.pdf}
        \label{fig:10-hourly-annual.pdf}
    \end{subfigure}
    \caption{
        System-level emission reduction: benchmark of procument policies and isolation of decarbonisation mechanisms. 
    \cref{fig:10-profile-volume.pdf} shows breakdown of emissions reduction in a local zone if 10\% of \gls{ci} load follows 24/7 hourly CFE procurement with 100\% target. The profile are volume effects are isolated.
    \cref{fig:10-hourly-annual.pdf} shows additional emissions reduction in a local zone by 100\% hourly over annual matching.}
    \label{fig:decarbonisation_story}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.95\columnwidth]{plots/10/2030/DE/p1/system_capacity_diff.pdf}
    \caption{
    Difference in generation capacity expansion in the whole European electricity system (including \gls{ci} resources) with and without clean energy procurement. 
    Selected scenario of Germany, 2030, technological palette~1.}
    \label{fig:10-2030-DE-p1-system_capacity_diff.pdf}
\end{figure}

>>>>

\subsection{Exploring mechanisms: how 24/7 CFE procurement decarbonises local grids}
\label{subsec:mechanisms}

\vspace{10pt}
\begin{res}
    24/7 CFE procurement reduces system-level \co \newline
    emissions through "profile" and "volume" mechanisms. 
\end{res}

\vspace{5pt}
\begin{res}
    The difference in impact that 24/7 CFE procurement and 100\% annual matching policies have on system decarbonisation amplifies with a cleaner state of local grids. %TODO REFINE
\end{res}

In the previous sections, we demonstrated that 24/7 clean energy procurement results in a substantial reduction of \co emissions in local electricity grids (Figures \ref{fig:10-2025-IE-p1-zone_emissions}, \ref{fig:10-2025-DE-p3-zone_emissions}, \ref{fig:10-2025-PL-p3-zone_emissions} and \ref{fig:10-2030-IE-p1-zone_emissions}).
In this section, we examine the underlying mechanisms that drive these observed reductions, so as to better understand how clean energy procurement and emissions reduction interrelate.

Two mechanisms are responsible for reducing system-wide emissions:

First, a \textit{profile} mechanism: participants doing 24/7 hourly matching procure clean energy resources that better match their demand patterns. 
As certain consumers timely align their demand with \gls{cfe} supply, the need for dispatchable generation that can firm intermittent renewable supply is lower in the rest of the system. 
This mechanism can reduce the utilisation of fossil-based generators, such as \gls{ocgt} power plants, which typically ramp up when wind and solar resources are scarce.

Second, a \textit{volume} mechanism: even the cost-optimal procurement strategy for 24/7 hourly matching might include some amount of excess clean energy.
If total \gls{cfe} generation of assets procured by 24/7 participants exceeds demand in a given hour and region, the “excess CFE” is not counted toward a CFE score; however, it is clean electricity that can potentially be stored (using batteries or LDES), or sold to the regional grid.
As for the latter, "excess CFE" might displace emitting grid generators or reduce the need to import electricity from neighbouring areas.

Several factors influence the degree to which profile and volume mechanisms contribute to decarbonising the local grids, such as the composition of the 24/7 portfolio, the volume of excess generation sold to the grid, and the electricity generation (and import) mix in the local zone.

In the following, we conduct an experiment to isolate the decarbonisation effects of the volume and profile mechanisms.
In the experiment, we reran the optimization with fixing an excess electricity sold to the grid to zero (NB excess is still possible, but can either be stored or curtailed). 
Such a model setup includes only a profile decarbonisation mechanism. 
We attribute the difference between emissions reduction in the experiment model setup and the base scenario to the volume mechanism.
As a result, the reduction of \co emissions in a local zone can be attributed either to excess electricity sold to the grid (volume effect) or to better alignment of CFE generation with demand (profile effect).

\cref{fig:10-profile-volume.pdf} illustrates the reduction in local zone emissions attributed to each of the two decarbonisation mechanisms.
In the plot, the x-axis shows the different states of electricity grids in which \gls{ci} consumers participate in 24/7 procurement.
The eight 'background grids' are generated from the combination of modelled wildcards: four zones and two years.

It can be seen in \cref{fig:10-profile-volume.pdf} that the overall impact of 24/7 CFE procurement is proportional to how clean the background grid on average is.
Thus, the largest impact is at 571.6~kgCO$_2$/a·MWh$^{-1}$ takes place in the Poland 2025 zone, and the lowest one is 3.7~kgCO$_2$/a·MWh$^{-1}$ in Denmark 2030 (NB 10\% of \gls{ci} sector participation rate with CFE 100\% score).
It is noteworthy that from 2025 to 2030, the decarbonisation effect of voluntary clean electricity procurement decreases over time in all zones since electricity grids become cleaner as a result of national climate policy programs and decomissioning of emitting power plants.
A comparison of the volume and profile mechanisms reveals that both contribute to the decarbonization of the local grids.
In particular, the profile mechanism contributes significantly to overall emissions reductions, with a major share across all zones regardless of their configuration.

Additional insights into decarbonisation mechanisms can be gained by examining the emission reductions achieved by 100\% renewables matching and 100\% 24/7 CFE matching policies.
\cref{fig:10-hourly-annual.pdf} compares the decarbonisation impact in a local zone of the two procurement policies in relative terms (percent point p.a.); the counterfactual scenario is a case when no \gls{ci} consumers participate in voluntary clean electricity procurement.
The figure shows that the relative impact of 24/7 CFE matching on system-level emissions \textit{increases over time}; this is despite the fact that the impact of procurement policies in absolute terms decreases over time, as shown above.
Furthermore, the self-cannibalisation effect of non-dispatchable generators, such as wind and solar, decreases the value of additional renewable capacity.
Thus, in the countries where governments commit to ambitious \gls{necp} targets, such as Ireland and Germany, 24/7 CFE matching has a stronger decarbonisation impact than 100\% annual matching with renewable energy. 
It is because the profile mechanism remains relevant even when renewable energy generators are self-cannibalizing; whereas the volume mechanism is amplified due to the diverse portfolio of procured resources, including batteries, long-term energy storage, and advanced clean firm technologies. As a result, clean electricity can be available during times when wind and solar generation is scarce system-wide.

\begin{figure}
    \centering
    \begin{subfigure}[t]{0.95\columnwidth}
        \centering
        \caption{Emissions reduction in a local zone with profile and volume mechanisms isolated. Data is normalised per MWh of demand participating in 24/7 procurement.}
        \includegraphics[width=0.95\columnwidth]{normalized_effects.pdf}
        \label{fig:10-profile-volume.pdf}
    \end{subfigure}
    \begin{subfigure}[t]{0.95\columnwidth}
        \centering
        \caption{Percentage of emissions reductions in a local zone as a result of a procurement policy compared with the no procurement policy case.}
        \includegraphics[width=0.95\columnwidth]{emissions_reduction_comparison.pdf}
        \label{fig:10-hourly-annual.pdf}
    \end{subfigure}
    \caption{
        System-level emission reduction: comparison of procurement policies and isolation of decarbonisation mechanisms.
        \cref{fig:10-profile-volume.pdf} shows breakdown of emissions reduction in a local zone if 10\% of \gls{ci} load follows 24/7 hourly procurement with CFE 100\% target. The profile are volume mechanisms are isolated.
        \cref{fig:10-hourly-annual.pdf} compares reductions in annual zone emissions in percent points achieved through 100\% annual renewable matching and 100\% 24/7 CFE matching; no procurement policy is assumed in the counterfactual scenario.}
    \label{fig:decarbonisation_story}
\end{figure}

% \subsection{System flexibility need reduction}
% \label{subsec:flexibility}

% \vspace{5pt}
% \begin{res}
%     24/7 CFE procurement reduces the needs for flexibility in the rest of the system.
% \end{res}

% The decarbonisation mechanisms discussed above suggest a displacement of fossil-fueled generators in the rest of the system in presense of voluntary clean electricity procurement.
% We illustrate this effect in \cref{fig:10-2030-DE-p1-system_capacity_diff.pdf}, which shows the difference in the generation capacity expansion of the whole European electricity system (excluding resources procured by participating customers) with and without commitments to clean electrity procurement (an illustraive scenario for Germany, 2030, technological palette~1). 
% Hourly CFE procurement benefits the system by having a pronounced effect of flexibility needs reduction -- the system requires less battery storage and less peaker capacity in a form of \gls{ocgt}. 
% In the selected example, 3.8~GW of load (10\% of \gls{ci} sectors in Germany) participate in the procurement.
% In a case of hourly matching with 100\% CFE target, this reduces \gls{ocgt} investments by 4.6~GW and battery storage by 2.6~GW in the rest of the system.
% The substitution of gas-fired capacity with clean resources procured by \gls{ci} consumers facilitates decarbonization of the entire system, as shown above.

% \begin{figure}
%     \centering
%     \includegraphics[width=0.95\columnwidth]{plots/10/2030/DE/p1/system_capacity_diff.pdf}
%     \caption{
%     Difference in generation capacity expansion in the whole European electricity system (including \gls{ci} resources) with and without clean energy procurement. 
%     Selected scenario of Germany, 2030, technological palette~1.}
%     \label{fig:10-2030-DE-p1-system_capacity_diff.pdf}
% \end{figure}
